Syntax and Semantics of Abstract Binding Trees

### 2015-11-11 paper.tex
Move definition of signature upwards; fix a few layout issues

\newcommand\MkSet[1]{\left\{#1\right\}}
\newcommand\Member[2]{\IMode{#1}\in\IMode{#2}}
\newcommand\EqMember[3]{\IMode{#1}=\IMode{#2}\in\IMode{#3}}
\newcommand\Of[2]{\IMode{#1}: \IMode{#2}}

@@

\newcommand\IsOperator[3]{\IMode{#1}\,{\Vdash}_{\scriptscriptstyle{\Operators}}\,\IMode{#2}: \OMode{#3}}

@@

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\to\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. By the Grothendieck
construction $\Iso{\Grothendieck{}{\Operators}}{\Comma{*}{\Operators}}$ we have
a category of elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are diagrams:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    \& *
      \arrow[dl, "\vartheta", swap]
      \arrow[dr, "\vartheta'"]
    \&
    \\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, "\Operators\Pair{\rho}{f}", swap]
    \&
    \& {\Operators\Pair{\Upsilon'}{a'}}
    \\
    {\Pair{\Upsilon}{a}}
      \arrow[rr, "\Pair{\rho}{f}", swap]
    \&
    \& {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\to\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  &{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}&&\,\mathrel{\ {{:}}}\,&&\Sets\\
  &{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}&&\,=\,\ && %
    \{\,\vartheta \in \Pair{\Upsilon}{a}%
      {{|}}\ {\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} = \Pair{\Upsilon}{a}%
  \}
\end{alignat*}

@@

For instance, consider the lambda calculus with a single sort, $\SortExp$; we
give its signature by asserting the following about its operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\lambda}{
    \MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}
These rules correspond to objects of $\Grothendieck{}{\Operators}$ and we
collect their fibers into sets of operators:
\begin{alignat*}{5}
{\pi_{\Operators}}^{-1}&\Pair{\Upsilon}{{\MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}}}\quad&&=
  &&\quad\{\,\lambda\,\}\\
{\pi_{\Operators}}^{-1}&\Pair{\Upsilon}{\MkArity{\MkValence{}{}{\SortExp},\MkValence{}{}{\SortExp}}{\SortExp}}\quad&&=
  &&\quad\{\,\mathtt{ap}\,\}\\
\bigcup_{\Pair{\Upsilon}{a}}&{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}\quad&&=
  &&\quad\{\,\lambda,\ \mathtt{ap}\,\}\\
\end{alignat*}
So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):

@@

\begin{definition}
  A \emph{signature} $\Sigma$ is a set of sorts $\Sorts$ together with a copresheaf $\Operators$ of operators.
\end{definition}

@@

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:\tau}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\to P_\tau}%
\]%
The interpretation of a variable
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ simply
projects it of the environment and embeds it into the model, as follows:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_3}"]
    \& {\IMode{V^{[\Gamma]}}} \arrow[r, "\IMode{\nu_{\Gamma}}"]
    \& {\IMode{P^{[\Gamma]}}} \arrow[r, "\IMode{\pi_x}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{var}
\]

Next we have the interpretation of metavariable applications
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
: \tau}_P$ where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$, which
proceeds by projecting the metavariable's interpretation from the environment
and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=7em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{S^{\{\Upsilon\}}}} \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{P^{[\vec{\tau}]}}}
  \end{tikzcd}
\end{gather*}

@@

Each argument
%
\begin{tikzcd}[ampersand replacement = \&]
  {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
    \arrow[r, "\IMode{\gamma_i}"]
  \& {\IMode{%
    P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
  }}
\end{tikzcd}
%
is the exponential transpose of the composite,
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\IMode{\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}}"]
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}}
      \arrow[r, "\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P}"]
    \& {\IMode{P_{\tau_i}}}
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}},\, \Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}}"]
    \& {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}},\, \Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}}"]
    \& {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}

>>>>

\newcommand\MkSet[1]{\left\{\,#1\,\right\}}
\newcommand\Member[2]{\IMode{#1}\in\IMode{#2}}
\newcommand\EqMember[3]{\IMode{#1}=\IMode{#2}\in\IMode{#3}}
\newcommand\Of[2]{\IMode{#1}: \IMode{#2}}

@@

\newcommand\IsOperator[3]{
  \IMode{#1}\,{\Vdash}_{\Sigma}\,\IMode{#2}: \OMode{#3}
}

@@

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\to\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. Together, $\Sorts$ and
$\Operators$ form a \emph{signature} $\Define{\Sigma}{\Pair{\Sorts}{\Operators}}$.

By the Grothendieck construction
$\Iso{\Grothendieck{}{\Operators}}{\Comma{*}{\Operators}}$ we have a category
of elements of $\Operators$ where objects are triples
$\Pair{\Pair{\Upsilon}{a}}{\vartheta}$, interpreted as
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$, and morphisms are diagrams:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    \& *
      \arrow[dl, "\vartheta", swap]
      \arrow[dr, "\vartheta'"]
    \&
    \\
    {\Operators\Pair{\Upsilon}{a}}
      \arrow[rr, "\Operators\Pair{\rho}{f}", swap]
    \&
    \& {\Operators\Pair{\Upsilon'}{a'}}
    \\
    {\Pair{\Upsilon}{a}}
      \arrow[rr, "\Pair{\rho}{f}", swap]
    \&
    \& {\Pair{\Upsilon'}{a'}}
  \end{tikzcd}
\]
We will write $\IsOperator{\Upsilon}{\vartheta}{a}$ in case
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and note that this judgment
enjoys the structural properties of weakening and exchange via functoriality of
$\Operators$. Operators $\vartheta_{0}, \ldots, \vartheta_{n}$ are defined by
specifying the fibers of
$\Of{\pi_{\Operators}}{\Grothendieck{}{\Operators}\to\SCtx\times\Discrete{\Arities}}$
in which they reside:
\begin{alignat*}{3}
  &{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}&&\,\mathrel{\ {{:}}}\,&&\Sets\\
  &{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}&&\,=\,\ && %
    \{\,\vartheta \in \Pair{\Upsilon}{a}%
      {{|}}\ {\pi_{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}} = \Pair{\Upsilon}{a}%
  \}
\end{alignat*}

@@

For instance, consider the lambda calculus with a single sort, $\SortExp$; we
give its signature by asserting the following about its operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\lambda}{
    \MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathtt{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}
These rules correspond to objects of $\Grothendieck{}{\Operators}$ and we
collect their fibers into sets of operators:
\begin{alignat*}{5}
{\pi_{\Operators}}^{-1}&\Pair{\Upsilon}{{\MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}}}\quad&&=
  &&\quad\MkSet{\lambda}\\
{\pi_{\Operators}}^{-1}&\Pair{\Upsilon}{\MkArity{\MkValence{}{}{\SortExp},\MkValence{}{}{\SortExp}}{\SortExp}}\quad&&=
  &&\quad\MkSet{\mathtt{ap}}\\
\bigcup_{\Pair{\Upsilon}{a}}&{{\pi_{\Operators}}^{-1}{\Pair{\Upsilon}{a}}}\quad&&=
  &&\quad\MkSet{\lambda,\ \mathtt{ap}}\\
\end{alignat*}
So far, we have made no use of symbols and parameters; however, consider the
extension of the calculus with assignables (references):

@@

@@

Then, the interpretation of a term in a model $P$ is a map from its environment
to $P$:
\[%
  \Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash%
      M:\tau}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\to P_\tau}%
\]

The interpretation of a variable
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ simply
projects it of the environment and embeds it into the model, as follows:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_3}"]
    \& {\IMode{V^{[\Gamma]}}} \arrow[r, "\IMode{\nu_{\Gamma}}"]
    \& {\IMode{P^{[\Gamma]}}} \arrow[r, "\IMode{\pi_x}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{var}
\]

Next interpret metavariable applications
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
: \tau}_P$ where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$, which
proceeds by projecting the metavariable's interpretation from the environment
and intantiating it via substitution:
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=7em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \& {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{metavar}
\]
where
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{S^{\{\Upsilon\}}}} \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{P^{[\vec{\tau}]}}}
  \end{tikzcd}
\end{gather*}

@@

Each argument
%
\begin{tikzcd}[ampersand replacement = \&]
  {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
    \arrow[r, "\IMode{\gamma_i}"]
  \& {\IMode{%
    P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
  }}
\end{tikzcd}
%
is the exponential transpose of the composite,
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=7.6em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\IMode{\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}}"]
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}}
      \arrow[r, "\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P}"]
    \& {\IMode{P_{\tau_i}}}
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}},\, \Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}}"]
    \& {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}},\, \Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}}"]
    \& {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}
